Nonparametric Goodness-of-Fit Tests for Discrete Null Distributions

### 2010-11-17 rjournal/RJtemplate.tex
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/ks-test@28 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%General idea of goodness-of-fit tests (hypothesis test about underlying distribution).
%Given an observed sequence of data, the typical question of statistical inference is to determine
%the underlying process from which it came. 
Goodness-of-fit tests are used to assess whether data are consistent
with a hypothesized null distribution.  The $\chi^2$ test is the best-known
parametric goodness-of-fit test, while the most popular nonparametric tests
are the classic test proposed by Kolmogorov and Smirnov followed closely by
several variants on an estimation procedure proposed by
Cram\'{e}r and von Mises (tests sometimes called Cram\'{e}r-von Mises-Smirnov
tests, or simply Cram\'{e}r-von Mises tests).

*** why "estimation procedure" above?
*** Citations in the paragraph above?  Also check the body of paper. ***

@@

*** Later, make sure we point out how using one of the ``old'' tests and
hoping for the best is a problem. ***

@@

The most popular nonparametric goodness-of-fit test is the
Kolmogorov-Smirnov test. The idea is fairly simple:
Given the cumulative distribution
function $F_0(x)$ of the continuous null distribution
and the empirical distribution function $F_{data}(x)$ of the observed data,  the test statistic is given by
\begin{align}
D = \sup_x \left| F_0(x)- F_{data}(x) \right|.
\end{align}
The distribution of $D$ does not depend on the hypothesized
distribution, making this a computationally
attractive method. \cite{slakter65} offers a standard presentation
of the test and its
performance relative to other algorithms. 
The test statistic (1) is easily adapted for one-sided tests.
For these, the absolute value is discarded and the tests are based
on either the supremum of the remaining difference (the `greater' testing
alternative) or by replacing the supremum with a negative infimum
(the `lesser' hypothesis alternative).  Tabulated p-values have been
available for these tests since (*** year ****).

**** check this above, both do away with the absolute value, I think? ***

The extension of the Kolmogorov-Smirnov test to non-continuous 
null distributions does not have such a clean solution. The formula of
the test statistic $D$ remains unchanged, but the distribution
of the test statistic is much more difficult to obtain. Unlike the 
continuous case, the null distribution depends on the null model.
It has been known since at least the 1950's that using the tables associated
with continuous null distributions results in conservative p-values
(**** citation?) when the null distribution is discontinuous 
(*** or should this be discrete? ***).  In the early 1970's, 
\citet{Conover1972} developed the method implemented here
for computing exact p-values
in the case of discrete null distributions.

@@

The test statistic is, in terms of the theory, the same as in the
continuous case; it would seem that no additional work would be required,
but unfortunately this is not the case.
Consider two non-decreasing functions $f$ and $g$, where the function $f$ is a step function with jumps on the set $\{x_1, \ldots x_N \}$ and $g$
is continuous (the classical Kolmogorov-Smirnov situation).
In order to determine the supremum of the difference between
these two functions, notice that
\begin{align}
\sup_x &\left| f(x)- g(x) \right|  \nonumber  \\
       &= \max_i \left( \left|g(x_i) - f(x_i) \right|, \lim_{x \rightarrow x_i} \left| g(x) - f(x_{i-1}) \right| \right) \\
&=  \max_i \bigg( \left|g(x_i) - f(x_i) \right|, \left| g(x_i) - f(x_{i-1}) \right| \bigg).
\end{align}
Computing the maximum over these $2N$ (*** why 2N? ***)
values (with $f$ equal to
$F_{data}(x)$ and $g$ equal to $F_0(x)$ as defined above) is clearly the 
most efficient way to compute the Kolmogorov-Smirnov test statistic for
a continuous null distribution. When the function $g$ is not
continuous, however, equality (3) does not hold in general because 
we cannot replace $\lim_{x\rightarrow x_e} g(x)$ with the value $g(x_i)$.

@@

Having obtained the test statistic, the p-value must then be calculated. 
For larger sample sizes, the null distribution coincides 
(***should this be
stated precisely as an asymptotic result???)
with the standard null distribution (CITATION???). 
When an exact p-value is required for
smaller sample sizes, the methodology in \citet{Conover1972} is used. 
Full details of the calculations are contained in source code
of our revised function \code{ks.test()} and in Conover's original paper.

** code hasn't previously existed?  I deleted this stuff, not sure if
it fits here.  Is it obvious (since we bothered to do this in the first
place)?  Or move it?

@@

While the Kolmogorov-Smirnov test may be the most popular of
the nonparametric goodness-of-fit tests, there is another family of 
tests which has been shown to be more powerful against a large class
of alternatives hypotheses. 
The original test was developed jointly (*** or was it simultaneously? ***) by
Harald Cram\'{e}r and Richard von Mises \citep{cramer1928, vonmises1928} and further adapted by \cite{anderson1952}, and  \cite{Watson1961}. 
The original test statistic, $W^2$, Anderson's $A^2$, and Watson's
$U^2$ are:
\begin{align}
W^2 &= n \cdot \int_{-\infty}^{\infty} \left[ F_{data}(x)- F_{0}(x) \right]^2 dF_0(x) \\
A^2 &= n \cdot \int_{-\infty}^{\infty} \frac{\left[F_{data}(x)- F_{0}(x) \right]^2}{F_0(x) -F_0(x)^2} dF_0(x) \\
U^2 &= n \cdot \int_{-\infty}^{\infty} \left[ F_{data}(x)- F_{0}(x) - W^2 \right]^2 dF_0(x)
\end{align}
As with the original Kolmogorov-Smirnov test statistic, these all have 
null distributions of their test statistics which are independent of the
specified null models.

The relative powers of these tests to different alternatives are
studied in depth in \cite{stephens1974}. In general, the $W^2$
statistic is recommended unless there is a compelling reason to
consider one of the others. The $A^2$ statistic was developed by
Anderson in the process of generalizing the test for the two-sample case,
but was shown to perform very similarly to the original statistic
in the one-sample setting (*** so why, then would we ever use it??? ***).
Watson's $U^2$ statistic was developed for distributions
which are cyclic (with an ordering to the support but
no natural starting point  *** how best to describe this? ***). 
For example, a distribution on the months of the year could be considered
cyclic. Watson's statistic is invariant to cyclical reordering (giving preference to one ordering would be unnatural).  **** is this last point necessary? ***

*** Is a second paragraph on Stephen's results really necessary?  Can we
pick and choose and get away with one paragraph? ***

@@

*** I edited, below, make sure you check it. ****

Generalizations of the Cram\'{e}r-von Mises tests to discrete
distributions were developed in \cite{choulakian1994}. As with the
Kolmogorov-Smirnov test, the forms of the test statistics are unchanged, although the discreteness allows for slightly simpler representations.
The null distribution of the test statistics are again hypothesis-dependent. The methods do not offer finite sample results, but rather show that the asymptotic null distribution is equal to a weighted sum of independent
chi-squared variables (with the weights depending on the particular
null distribution). (*** is this true for all 3? ***)
We implement this asymptotic distribution here;
the original papers (*** really, all of them? ***)
show that this approximation is conservative 
and asymptotically equivalent to the true null distribution.

*** What does it mean for a distribution to be equal to a weighted sum of
independent chi-squared variables?  I think this needs to be stated more
precisely, above.  A formal definition of $Q$ is needed someplace, too,
as I'll probably note below. ***

@@

There is no analytic solution to the integral, so the integration is
accomplished numerically. This seems fine in most situations we considered,
but numerical issues appear in the regime of large test statistics $x$
(or, equivalently, small p-values).
The function $\theta(\cdot, x)$ is linear in $x$; as the test statistic 
grows the corresponding periodicity of the integrand decreases and
the approximation becomes unstable. 
%This is further magnified in situations where the 
%$p$-values should be very small, where tiny fluctuations (undetectable
%elsewhere) become quite prominent. 
We resolve this problem by using a simple conservative approximation to
avoid the numerical instability. Consider the
following inequality:
\begin{align}
\mathbb{P} \left(\sum_{i=1}^{p} \lambda_i \chi^2_1 \geq x \right) &\leq \mathbb{P} \left( \lambda_{max} \sum_{i=1}^{p} \chi^2_1 \geq x \right) \\
&= \mathbb{P} \left(\chi^2_p \geq \frac{x}{p \, \lambda_{max}} \right)
\end{align}
The values for the weighted sum can be bounded using a simple transformation
and a chi-squared distribution of a higher degree of freedom. 
The original formulation is preferable for most p-values, while
this approximation is useful for smaller p-values (smaller than 0.001,
based on our observations of the numerical instablity of the original
formulaton); it should not greatly affect the interpretation of the results.

*** How is the cutoff established in the code? ***

Figure \ref{cvmissues} shows the non-monotonicity of the function 
(*** which function, reference it? ***) as the test statistic grows, 
compared to the conservative approximation.

*** stuff deleted, redundant ***.

\begin{figure}
\begin{center}
\includegraphics[scale=0.4]{fig1.pdf}
\end{center}
\caption{Plot of p-value for given test statistics using numerical integration
(pink) versus a conservative chi-squared bound (dashed blue). }
\label{cvmissues}
\end{figure}

@@

*** Discussion of what we might have broken: cases where the user
provided a discrete distribution to the original ks.test() even though
it wasn't intended to be supported. ***

@@

TAYLOR: good cyclical example, relating to new material about to go into
earlier section.

@@

*** I commented out a bunch of stuff.  I think the discussion can be short
and sweet. ***

@@

*** Hmm... if you claim the following, you want to make sure it is true.
Did you read it somewhere? ***

@@

In the continuous setting, both of 
the Kolmogorov-Smirnov and the Cram\'{e}r-von Mises tests have two-sample analogues. Here data are
observed from two processes, and the hypothesis tested is 
whether they came from the same (but unspecified) distribution. There
does not exist an analogous theory for discrete distributions. 
This comes from the fact that the discrete null distributions
of the test statistics depend on the exact null distribution; 
therefore the two-sample case would surely have to depend
on the exact distributions used as well (*** is this true? ***), 
which are generally not even stated in the two-sample case.

>>>>

%General idea of goodness-of-fit tests (hypothesis test about underlying distribution).
%Given an observed sequence of data, the typical question of statistical inference is to determine
%the underlying process from which it came. 
Goodness-of-fit tests are used to assess whether data are consistent
with a hypothesized null distribution.  The $\chi^2$ test is the best-known
parametric goodness-of-fit test, while the most popular nonparametric tests
are the classic test proposed by Kolmogorov and Smirnov followed closely by
several variants on a procedure proposed by
Cram\'{e}r and von Mises (tests sometimes called Cram\'{e}r-von Mises-Smirnov
tests, or simply Cram\'{e}r-von Mises tests) .

%*** why "estimation procedure" above?
%*** Citations in the paragraph above?  Also check the body of paper. ***

@@

%*** Later, make sure we point out how using one of the ``old'' tests and
%hoping for the best is a problem. ***

@@

The most popular nonparametric goodness-of-fit test is the
Kolmogorov-Smirnov test. The idea is fairly simple:
Given the cumulative distribution
function $F_0(x)$ of the continuous null distribution
and the empirical distribution function $F_{data}(x)$ of the observed data,  the test statistic is given by
\begin{align}
D = \sup_x \left| F_0(x)- F_{data}(x) \right|.
\end{align}
The distribution of $D$ does not depend on the hypothesized
distribution, making this a computationally
attractive method. \cite{slakter65} offers a standard presentation
of the test and its
performance relative to other algorithms. 
The test statistic (1) is easily adapted for one-sided tests.
For these, the absolute value is discarded and the tests are based
on either the supremum of the remaining difference (the `greater' testing
alternative) or by replacing the supremum with a negative infimum
(the `lesser' hypothesis alternative).  Tabulated p-values have been
available for these tests since 1933. %(*** year ****).

%**** check this above, both do away with the absolute value, I think? ***

The extension of the Kolmogorov-Smirnov test to non-continuous 
null distributions does not have such a clean solution. The formula of
the test statistic $D$ remains unchanged, but the distribution
of the test statistic is much more difficult to obtain. Unlike the 
continuous case, the null distribution depends on the null model.
It has been known since at least the 1950's that using the tables associated
with continuous null distributions results in conservative p-values
when the null distribution is discontinuous 
\citep{stephens1974}.  In the early 1970's, 
\citet{Conover1972} developed the method implemented here
for computing exact p-values
in the case of discrete null distributions.

@@

The test statistic is, in terms of the theory, the same as in the
continuous case; it would seem that no additional work would be required,
but unfortunately this is not the case.
Consider two non-decreasing functions $f$ and $g$, where the function $f$ is a step function with jumps on the set $\{x_1, \ldots x_N \}$ and $g$
is continuous (the classical Kolmogorov-Smirnov situation).
In order to determine the supremum of the difference between
these two functions, notice that
\begin{align*}
&\sup_x \left| f(x)- g(x) \right|  \nonumber  \\
       &= \max_i \left( \max\left( \left|g(x_i) - f(x_i) \right|, \lim_{x \rightarrow x_i} \left| g(x) - f(x_{i-1}) \right| \right)\right) \\
&=  \max_i \bigg(\max \bigg( \left|g(x_i) - f(x_i) \right|, \left| g(x_i) - f(x_{i-1}) \right| \bigg) \bigg).
\end{align*}
Computing the maximum over these $2N$
values (with $f$ equal to
$F_{data}(x)$ and $g$ equal to $F_0(x)$ as defined above) is clearly the 
most efficient way to compute the Kolmogorov-Smirnov test statistic for
a continuous null distribution. When the function $g$ is not
continuous, however, equality (3) does not hold in general because 
we cannot replace $\lim_{x\rightarrow x_e} g(x)$ with the value $g(x_i)$.

@@

Having obtained the test statistic, the p-value must then be calculated. 
For larger sample sizes, the null distribution can be approximated
with the standard null distribution. 
When an exact p-value is required for
smaller sample sizes, the methodology in \citet{Conover1972} is used. 
Full details of the calculations are contained in source code
of our revised function \code{ks.test()} and in Conover's original paper.

%** code hasn't previously existed?  I deleted this stuff, not sure if
%it fits here.  Is it obvious (since we bothered to do this in the first
%place)?  Or move it?

@@

While the Kolmogorov-Smirnov test may be the most popular of
the nonparametric goodness-of-fit tests, there is another family of 
tests which has been shown to be more powerful against a large class
of alternatives hypotheses. 
The original test was developed by
Harald Cram\'{e}r and Richard von Mises \citep{cramer1928, vonmises1928} and further adapted by \cite{anderson1952}, and  \cite{Watson1961}. 
The original test statistic, $W^2$, Anderson's $A^2$, and Watson's
$U^2$ are:
\begin{align}
W^2 &= n \cdot \int_{-\infty}^{\infty} \left[ F_{data}(x)- F_{0}(x) \right]^2 dF_0(x) \\
A^2 &= n \cdot \int_{-\infty}^{\infty} \frac{\left[F_{data}(x)- F_{0}(x) \right]^2}{F_0(x) -F_0(x)^2} dF_0(x) \\
U^2 &= n \cdot \int_{-\infty}^{\infty} \left[ F_{data}(x)- F_{0}(x) - W^2 \right]^2 dF_0(x)
\end{align}
As with the original Kolmogorov-Smirnov test statistic, these all have 
null distributions of their test statistics which are independent of the
specified null models.

The relative powers of these tests to different alternatives are
studied in depth in \cite{stephens1974}. The $W^2$
statistic was the original test statistic; it has the benefit of having
the simplest form.
The $A^2$ statistic was developed by
Anderson in the process of generalizing the test for the two-sample case.
Watson's $U^2$ statistic was developed for distributions
which are cyclic (with an ordering to the support but
no natural starting point), as it is invariant to cyclic reordering of
the support. 
For example, a distribution on the months of the year could be considered
cyclic.

@@

%*** I edited, below, make sure you check it. ****

Generalizations of the Cram\'{e}r-von Mises tests to discrete
distributions were developed in \cite{choulakian1994}. As with the
Kolmogorov-Smirnov test, the forms of the test statistics are unchanged, although the discreteness allows for slightly simpler representations.
The null distribution of the test statistics are again hypothesis-dependent. The methods do not offer finite sample results, but rather 
show that the asymptotic distribution of the test statistic under the null hypothesis is equal to the distribution of a weighted sum of independent
chi-squared variables (with the weights depending on the particular
null distribution). 
We implement this asymptotic distribution here;
the original papers %(*** really, all of them? ***)
show that this approximation is conservative 
and asymptotically equivalent to the true null distribution.

%*** What does it mean for a distribution to be equal to a weighted sum of
%independent chi-squared variables?  I think this needs to be stated more
%precisely, above.  A formal definition of $Q$ is needed someplace, too,
%as I'll probably note below. ***

@@

There is no analytic solution to the integral, so the integration is
accomplished numerically. This seems fine in most situations we considered,
but numerical issues appear in the regime of large test statistics $x$
(or, equivalently, small p-values).
The function $\theta(\cdot, x)$ is linear in $x$; as the test statistic 
grows the corresponding periodicity of the integrand decreases and
the approximation becomes unstable. 
%This is further magnified in situations where the 
%$p$-values should be very small, where tiny fluctuations (undetectable
%elsewhere) become quite prominent. 
We resolve this problem by using a simple conservative approximation to
avoid the numerical instability. Consider the
following inequality:
\begin{align}
\mathbb{P} \left(\sum_{i=1}^{p} \lambda_i \chi^2_1 \geq x \right) &\leq \mathbb{P} \left( \lambda_{max} \sum_{i=1}^{p} \chi^2_1 \geq x \right) \\
&= \mathbb{P} \left(\chi^2_p \geq \frac{x}{p \, \lambda_{max}} \right)
\end{align}
The values for the weighted sum can be bounded using a simple transformation
and a chi-squared distribution of a higher degree of freedom. 
The original formulation is preferable for most p-values, while
this approximation is useful for smaller p-values (smaller than 0.001,
based on our observations of the numerical instability of the original
formulation); it should not greatly affect the interpretation of the results.

%*** How is the cutoff established in the code? ***

Figure \ref{cvmissues} shows the non-monotonicity of the numerical
evaluation of equation 7 for the null distribution which is
uniform on the set $\{1,2,3\}$, as well as the behavior of the
conservative approximation in equation 9.

\begin{figure}
\begin{center}
\includegraphics[scale=0.4]{fig1.pdf}
\end{center}
\caption{Plot of p-value for given test statistics using numerical integration
(pink) versus a conservative chi-squared bound (dashed blue). The null distribution
is unform on the set $\{1,2,3\}$. }
\label{cvmissues}
\end{figure}

@@

%*** Discussion of what we might have broken: cases where the user
%provided a discrete distribution to the original ks.test() even though
%it wasn't intended to be supported. ***

@@

Cyclical example:

\begin{Schunk}
\begin{Sinput}
> library(cvm.test)
> x <- sample(1:4, 20, replace=T)
> cvm.test(x, ecdf(1:4), type='U2')
\end{Sinput}
\begin{Soutput}
	Cramer-von Mises - U2

data:  x 
U2 = 0.075, p-value = 0.4027
alternative hypothesis: Two.sided 
\end{Soutput}
\begin{Sinput}
> y <- x%%4 + 1
> cvm.test(y, ecdf(1:4), type = 'U2')
\end{Sinput}
\begin{Soutput}
	Cramer-von Mises - U2

data:  y 
U2 = 0.075, p-value = 0.4027
alternative hypothesis: Two.sided 
\end{Soutput}
\end{Schunk}

@@

%*** I commented out a bunch of stuff.  I think the discussion can be short
%and sweet. ***

@@

%*** Hmm... if you claim the following, you want to make sure it is true.
%Did you read it somewhere? ***

@@

In the continuous setting, both of 
the Kolmogorov-Smirnov and the Cram\'{e}r-von Mises tests have two-sample analogues. Here data are
observed from two processes, and the hypothesis tested is 
whether they came from the same (but unspecified) distribution. There
does not exist an analogous theory for discrete distributions. 
This comes from the fact that the discrete null distributions
of the test statistics depend on the exact null distribution; 
therefore the two-sample case would surely have to depend
on the exact distributions used as well %(*** is this true? ***), 
which are generally not even stated in the two-sample case.
